---
title: "Regression Models Course Project"
author: "Carlos Saquel"
date: "December 06, 2018"
output: pdf_document
---
# Executive Summary

Motor Trend, an automobile trend magazine is interested in exploring the relationship between a set of variables and miles per gallon (MPG) outcome. In this project, we will analyze the mtcars dataset from the 1974 Motor Trend US magazine to answer the following questions:

- Is an automatic or manual transmission better for miles per gallon (MPG)?

- How different is the MPG between automatic and manual transmissions?


#Data Description

We analyze the **mtcars** data set through Regression Modelling and exploratory analysis to show how automatic (am = 0) and manual (am = 1) transmissions features affect the MPG feature.

The data was extracted from the 1974 Motor Trend US magazine, which comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

The data set consists of a data frame with 32 observations (nrow) and 11 variables (ncol).

- **mpg**: Miles per US gallon
- **cyl**: Number of cylinders
- **disp**: Displacement (*cubic inches*)
- **hp**: Gross horsepower
- **drat**: Rear axle ratio
- **wt**: Weight (1000 lb)
- **qsec**: 1/4 mile time
- **vs**: Engine (0 = *V-shaped*, 1 = *straight*)
- **am**: Transmission (0 = *automatic*, 1 = *manual*)
- **gear**: Number of forward gears
- **carb**: Number of carburetors

```{r load, echo = FALSE, message=FALSE, comment=""}
library(ggplot2)
library(dplyr)
library(knitr)
data(mtcars)
```


# Appendix

## A.1 - Exploratory Data Analysis

To begin the analysis we review the data taking a sample and seeing how it is structured (table 1, 2 and 3).

```{r table1, echo = FALSE, message=FALSE, comment=""}
data <- mtcars
colnames(data) <- paste0("**",names(data),"**")
rownames(data) <- paste0("**",rownames(data),"**")
knitr::kable(head(data,6), align = "c", caption = "Sample of data set mtcars")
```

```{r plot1, echo = FALSE, message=FALSE, fig.cap="Plot mtcars variables"}
pairs(mtcars)
#library(GGally)
#ggpairs(mtcars, aes(alpha = 0.4))
```

```{r table2, echo = FALSE, message=FALSE, comment=""}
a<-list()
for (i in 1:dim(mtcars)[2]){
      a[i]<-class(mtcars[,i])
}
c<- data.frame(t(a))
colnames(c) <- paste0("**",names(mtcars),"**")
#rownames(c) <- "**class**"
knitr::kable(c, align = "c", pading = 3, caption = "Class of the data set mtcars")
```





Vamos a ajustar algunas las variables xxxxxxxx que son numericas pero deberian ser tipo factor.

```{r table3, echo = FALSE, message=FALSE, comment=""}
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)

knitr::kable(summary(mtcars[,1:5]), align = "c", caption = "Structure of the data set mtcars"); kable(summary(mtcars[,6:11]), align = "c", caption = "")
```

