---
title: "Regression Models Course Project"
author: "Carlos Saquel"
date: "December 08, 2018"
output: pdf_document
---
# Executive Summary

Motor Trend, an automobile trend magazine is interested in exploring the relationship between a set of variables and miles per gallon (MPG) outcome. In this project, we will analyze the mtcars dataset from the 1974 Motor Trend US magazine to answer the following questions:

- Is an automatic or manual transmission better for miles per gallon (MPG)?

- How different is the MPG between automatic and manual transmissions?


#Data Description

We analyze the **mtcars** data set through Regression Modelling and exploratory analysis to show how automatic (*am* = 0) and manual (*am* = 1) transmissions features affect the MPG feature.

The data was extracted from the 1974 Motor Trend US magazine, which comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

The data set consists of a data frame with 32 observations (nrow) and 11 variables (ncol).

- **mpg**: Miles per US gallon
- **cyl**: Number of cylinders
- **disp**: Displacement (*cubic inches*)
- **hp**: Gross horsepower
- **drat**: Rear axle ratio
- **wt**: Weight (1000 lb)
- **qsec**: 1/4 mile time
- **vs**: Engine (0 = *V-shaped*, 1 = *straight*)
- **am**: Transmission (0 = *automatic*, 1 = *manual*)
- **gear**: Number of forward gears
- **carb**: Number of carburetors

```{r load, echo = FALSE, message=FALSE, comment=""}
library(ggplot2); library(dplyr); library(knitr); library(MASS); data(mtcars)
data <- mtcars
```

```{r table3, echo = FALSE, eval=FALSE, message=FALSE, comment=""}
mtcars$cyl <- factor(mtcars$cyl)
mtcars$am <- factor(mtcars$am,labels=c("Automatic","Manual"))
mtcars$vs <- factor(mtcars$vs,labels=c("V-shaped","straight"))
mtcars$gear <- factor(mtcars$gear); mtcars$carb <- factor(mtcars$carb)
```

```{r table1, echo = FALSE, eval=FALSE, message=FALSE, comment=""}
knitr::kable(head(data,5), align = "c", caption = "Sample of data set mtcars"); 
a<-list();for (i in 1:dim(data)[2]){a[i]<-class(data[,i])}
knitr::kable(data.frame(t(a)), align = "c", pading = 3, 
             caption = "Class of the data set mtcars")
```


```{r plot1, eval=FALSE, echo=FALSE, message=FALSE}
p <- ggplot(mtcars, aes(x=factor(am,labels=c("Automatic","Manual")), y=mpg, 
      fill = factor(am,labels=c("Automatic","Manual")))) + geom_boxplot() +
      geom_dotplot(binaxis='y', stackdir='center', dotsize=1, fill ="black") +
      scale_fill_discrete(name = "Transmission") + xlab("Transmission") + 
      ylab("Miles per US gallon") + labs(caption = 
      "Figure 1: Miles per US gallon v/s Transmission (Automatic/Manual)");p
```


# Exploratory Data Analysis

First we must determine whether or not there really is a difference between automatic and manual transmissions in terms of mpg. For this purpose, a review of the data is made (see *Appendix A.1*)

In the **Figure1** in *Appendix A.1* the boxplot with the information of the mileage vs the transmission is shown. In this diagram you can clearly see that there is a difference in the mileage (*mpg*) depending on the type of transmission (automatic or manual). Where the median mileage (*mpg*) for an automatic transmission is `17.3` and for a manual transmission is `22.8`.


# Multivariable Regression Model

For the analysis we will consider 3 models (see *Appendix A.2*).

- **fit_base**: Regression model between `mpg` and` am`.

- **fit_best**: Regression model between `mpg` the variables determined by the` stepAIC` function.

- **fit_full**: Regression model between `mpg` and all the data set variables.


Once the `fit_best` model has been obtained (*mpg ~ cyl + hp + wt + am*) we will use the` anova` function to determine which of these three models best fits the data with the least possible complexity.

```{r fit 0, message=FALSE, include=FALSE}
fit_base <- lm(mpg ~ am, data = mtcars)
fit_full <- lm(mpg ~ ., data = mtcars)
fit_best <- stepAIC(fit_full, direction = "both")
```

```{r anova, message=FALSE, include=FALSE}
anova(fit_base, fit_best)
```

According to the analysis of variance (ver *Apendix A.2*) the p-value is significant, hence we reject the null hypothesis that the variables *cyl*, *hp* and *wt* do not contribute to the accuracy of the model.
We double-check the residuals for non-normality (*Apendix A.2 - Figure 2*) and can see they are all normally distributed and homoskedastic.


```{r fit 0a,echo=FALSE, message=FALSE, eval=TRUE}
d<-data.frame()
d[1,1] <- round(summary(fit_base)$r.squared,4); d[1,2] <- round(summary(fit_best)$r.squared,4)
colnames(d) <- c("fit_base", "fit_best"); rownames(d) <- "R^2^"
knitr::kable(d, align = "c", pading = 3, caption = "Comparacion R^2^ entre modelos")
```

The model `fit_best` explains 84.96% of the variance and as a result, *cyl*, *hp*, *wt*  did affect the correlation between *mpg* and *am*. Thus, we can say the difference between automatic and manual transmissions is 1.81 MPG.


# Conclusions

Based on the observations of our best fit model, we can conclude:

Considering the summaries of the models, it is observed that the model *fit_best* has a significant difference in the parameters with *fit_base*. For *fit_base*, cars with manual transmission get more miles per gallon compared to automobiles with automatic transmission. (7.25 *mpg* set by *hp*, *cyl*, and *wt*), in the case of *fit_best* is 1.81 in the same conditions.

The *mpg* will decrease by 2.5 (adjusted by *hp*, *cyl* and *am*) for every 1000 pounds of weight. *mpg* decreases negligibly with the increase of *hp*.

If the number of cylinders, *cil* increases from 4 to 6 and 8, *mpg* will decrease by a factor of 3 and 2.2 respectively (set by *hp*, *wt* and *am*).


# Appendix

## A.1 - Data Processing

First we load the data and the libraries.

```{r dProc0, ref.label='load', echo = TRUE, eval=TRUE, message=FALSE, comment=""}
```


Then we start with the analysis, review the data taking a sample and see how it is structured (table 2, 3).

```{r dProc1, ref.label='table1', echo = TRUE, eval=TRUE, message=FALSE, comment=""}
```

```{r dProc2, ref.label='table2', echo = TRUE, eval=TRUE, message=FALSE, comment=""}
```


Observing the data, let's adjust the variables **cyl**, **hp**, **am**, **gear** and **carb** which are numeric but should be factor type.

```{r dProc3, ref.label='table3', echo = TRUE, eval=TRUE, message=FALSE, comment=""}
```

With this we can compare how the type of transmission (automatic - manual) affects the mileage (*mpg*).

```{r dProc4, eval=TRUE, echo=TRUE, message=FALSE, comment="", ref.label='plot1'}
```


## A.2 - Multivariable Regression Model and inferences.
**Next, 3 linear models are generated:**

```{r fit 1,echo=TRUE, message=FALSE, eval=TRUE, comment=""}
fit_base <- lm(mpg ~ am, data = mtcars)
fit_full <- lm(mpg ~ ., data = mtcars)
fit_best <- stepAIC(fit_full, direction = "both")
```


**Analysis of variance of the models.**


```{r fit 2,echo=TRUE, message=FALSE, eval=TRUE, comment=""}
anova(fit_base, fit_best)
```

**Summary *fit_base* and *fit_best*.**
```{r fit 3,echo=TRUE, message=FALSE, eval=TRUE, comment=""}
summary(fit_base)
summary(fit_best)
```

**Check residuals**

```{r plot 3,echo=TRUE, message=FALSE, eval=TRUE}
par(mfrow=c(2,2), oma=c(2,0,0,0))
plot(fit_best)
mtext(side=1,line=0,cex=1, outer=T,"Figure 2: Check residuals")
```


